---
title: "ENV797 - Forecasting Competition: First Submission"
author: "Joy Wu, Aisha Shen"
output: pdf_document
always_allow_html: true
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: sentence
---

## Setting R code chunk options

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

## Loading packages

```{r package, message=FALSE, warning=FALSE}
library(lubridate)
library(ggplot2)
library(forecast)  
library(Kendall)
library(tseries)
library(outliers)
library(tidyverse)
library(smooth)
library(zoo)
library(kableExtra)
library(readxl)
library(tidymodels)
library(modeltime)
library(timetk)
library(tune)
library(dials)

```

## Importing data

The dataset comes from the Kaggle competition platform. There are three files: hourly electricity demand (load), hourly temperature, and hourly relative humidity from January 2005 to June 2011. The load file has one row per day with 24 columns (h1 to h24) representing hourly readings. The temperature and humidity files have one row per hour with readings from 28 weather stations.

```{r}
#Importing data from excel files
load_data <- read_excel("./Data/load.xlsx")
temp_data <- read_excel("./Data/temperature.xlsx")
rh_data   <- read_excel("./Data/relative_humidity.xlsx")

#Inspect load data
head(load_data)
nobs <- nrow(load_data)
summary(load_data[,3:26]) #summary of hourly columns only
```
## Wrangling and Processing the Dataset

We need to transform the hourly data into daily data by taking the average of the 24-hour data points. The load file already has one row per day with columns h1 through h24, so we average across columns. For temperature and humidity, we aggregate hourly observations to daily means using pipes.

### Processing load data
```{r}
load_daily <- load_data %>%
  mutate(date = ymd(date)) %>%
  rename(Date = date) %>%
  pivot_longer(cols = h1:h24, 
               names_to = "hour", 
               values_to = "load") %>%
  group_by(Date) %>%
  summarise(daily_mean_load = mean(load, na.rm = TRUE)) 

summary(load_daily)

#Check for NAs in our processed daily load
cat("Number of NAs in daily mean load:", 
    sum(is.na(load_daily$daily_mean_load)), "\n")
```
### Processing temperature data

```{r}
temp_daily <- temp_data %>%
  mutate(date = ymd(date)) %>%
  rename(Date = date) %>%
  pivot_longer(cols = t_ws1:t_ws28, 
               names_to = "station", 
               values_to = "temp") %>%
  group_by(Date) %>%
  summarise(daily_mean_temp = mean(temp, na.rm = TRUE)) %>%
  filter(!is.na(Date))

temp_daily_allsites <- temp_data %>%
  mutate(date = ymd(date)) %>%
  rename(Date = date) %>%
  filter(!is.na(Date)) %>%        
  group_by(Date) %>%
  summarise(across(t_ws1:t_ws28, ~ mean(.x, na.rm = TRUE)))

summary(temp_daily)

```
### Processing relative humidity data
```{r}
rh_daily <- rh_data %>%
  mutate(date = ymd(date)) %>%
  rename(Date = date) %>%
  pivot_longer(cols = rh_ws1:rh_ws28, 
               names_to = "station", 
               values_to = "rh") %>%
  group_by(Date) %>%
  summarise(daily_mean_rh = mean(rh, na.rm = TRUE)) %>%
  filter(!is.na(Date))
summary(rh_daily)

rh_daily_allsites <- rh_data %>%
  mutate(date = ymd(date)) %>%
  rename(Date = date) %>%
  filter(!is.na(Date)) %>%        
  group_by(Date) %>%
  summarise(across(rh_ws1:rh_ws28, ~ mean(.x, na.rm = TRUE)))

```
### Combining all daily data into one data frame
```{r}
#Join load, temperature, and humidity into one data frame by date
daily_data <- load_daily %>%
  left_join(temp_daily, by = "Date") %>%
  left_join(rh_daily, by = "Date")

head(daily_data)
summary(daily_data)
```

## Initial Plots
```{r}
#Plot daily load over time
ggplot(daily_data, aes(x = Date, y = daily_mean_load)) +
  geom_line(color = "blue") +
  ylab("Daily Mean Load (Wh)") +
  xlab("Date") +
  ggtitle("Daily Electricity Demand: Jan 2005 - Jun 2011")

#Plot daily temperature over time
ggplot(daily_data, aes(x = Date, y = daily_mean_temp)) +
  geom_line(color = "red") +
  ylab("Daily Mean Temperature (F)") +
  xlab("Date") +
  ggtitle("Daily Temperature: Jan 2005 - Jun 2011")

#Plot daily relative humidity over time
ggplot(daily_data, aes(x = Date, y = daily_mean_rh)) +
  geom_line(color = "green") +
  ylab("Daily Mean Relative Humidity (%)") +
  xlab("Date") +
  ggtitle("Daily Relative Humidity: Jan 2005 - Jun 2011")
```

## Creating Time Series Objects
Since this is daily data with more than one seasonal component (weekly and yearly patterns), we use `msts()` instead of `ts()`. The seasonal periods are 7 (weekly cycle) and 365.25 (yearly cycle, accounting for leap years).
```{r}
#Create msts object for daily load
ts_daily_load <- msts(daily_data$daily_mean_load,
                      seasonal.periods = c(7, 365.25),
                      start = c(2005, 1, 1))

#Create msts object for daily temperature (for potential future use)
ts_daily_temp <- msts(daily_data$daily_mean_temp,
                      seasonal.periods = c(7, 365.25),
                      start = c(2005, 1, 1))

#Create msts object for daily relative humidity (for potential future use)
ts_daily_rh <- msts(daily_data$daily_mean_rh,
                    seasonal.periods = c(7, 365.25),
                    start = c(2005, 1, 1))

```
## Decomposing the Time Series

Using `mstl()` from package `forecast` to decompose the multi-seasonal time series. 
```{r}
#Decompose daily load series using mstl
ts_daily_load %>% mstl() %>%
  autoplot() +
  ggtitle("Decomposition of Daily Electricity Demand")
```
## ACF and PACF
```{r}
#ACF and PACF plots to inspect autocorrelation structure
par(mfrow=c(1,2))
Acf(daily_data$daily_mean_load, lag.max = 40, 
    main = "ACF - Daily Load")
Pacf(daily_data$daily_mean_load, lag.max = 40, 
     main = "PACF - Daily Load")
par(mfrow=c(1,1))
```

## Setting Up Training and Validation Sets

Training data spans January 1, 2005 to June 30, 2010. The validation set is July 1, 2010 to July 31, 2010 (31 days). After selecting the best model, we retrain on the full dataset (January 2005 - June 2011) and forecast July 2011.
```{r}
#Count how many days fall in the training period (Jan 2005 - Jun 2010)
n_train <- sum(daily_data$Date <= as.Date("2010-06-30"))

#Count how many days fall in the validation period (Jul 2010)
n_val <- sum(daily_data$Date >= as.Date("2010-07-01") & 
             daily_data$Date <= as.Date("2010-07-31"))

#Extract the training portion from the full time series
ts_load_train <- subset(ts_daily_load, end = n_train)

#Extract the validation portion
ts_load_val <- subset(ts_daily_load, 
                      start = n_train + 1, 
                      end = n_train + n_val)

#Plot both sets
autoplot(ts_load_train) +
  ylab("Daily Mean Load") +
  ggtitle("Training Data: Jan 2005 - Jun 2010")

autoplot(ts_load_val) +
  ylab("Daily Mean Load") +
  ggtitle("Validation Data: Jul 2010")
```
## Fitting Models and Forecasting
### Model: Arithematic Mean
```{r}
# Fit mean model on training data
mean_fit <- meanf(ts_load_train, h = n_val)

# Plot forecast
autoplot(mean_fit) +
  ylab("Daily Mean Load") +
  ggtitle("Arithmetic Mean Forecast (Validation)")

# Overlay forecast with actual validation data
autoplot(ts_load_val) +
  autolayer(mean_fit, series = "Mean Forecast", PI = FALSE) +
  ylab("Daily Mean Load") +
  ggtitle("Mean Forecast vs Observed: Jul 2010")

# Residual Analysis
checkresiduals(mean_fit)

# Q-Q Plot for Mean Model
qqnorm(residuals(mean_fit))
qqline(residuals(mean_fit), col = "red")
```


### Model: Naive
```{r}
# Fit naive model on training data
naive_fit <- naive(ts_load_train, h = n_val)

# Plot forecast
autoplot(naive_fit) +
  ylab("Daily Mean Load") +
  ggtitle("Naive Forecast (Validation)")

# Overlay forecast with actual validation data
autoplot(ts_load_val) +
  autolayer(naive_fit, series = "Naive Forecast", PI = FALSE) +
  ylab("Daily Mean Load") +
  ggtitle("Naive vs Observed: Jul 2010")

# Residual Analysis
checkresiduals(naive_fit)

# Q-Q Plot for Naive Model
qqnorm(residuals(naive_fit))
qqline(residuals(naive_fit), col = "red")
```

### Model: Seasonal Naive
```{r}
#Fit seasonal naive model on training data
snaive_fit <- snaive(ts_load_train, h = n_val)

#Plot forecast
autoplot(snaive_fit) +
  ylab("Daily Mean Load") +
  ggtitle("Seasonal Naive Forecast for Jul 2010 (Validation)")

#Overlay forecast with actual validation data
autoplot(ts_load_val) +
  autolayer(snaive_fit, series = "Seasonal Naive", PI = FALSE) +
  ylab("Daily Mean Load") +
  ggtitle("Seasonal Naive vs Observed: Jul 2010")
```


## Checking Residuals

Following M8, we use `checkresiduals()` to evaluate the model fit. We want residuals that resemble white noise: no patterns, centered around zero, normally distributed.
```{r}
#Check residuals for Seasonal Naive
checkresiduals(snaive_fit)
```
## Checking Forecast Accuracy Score

```{r}
snaive_scores <- forecast::accuracy(snaive_fit, ts_load_val)
snaive_scores
```


## Enhancing Model robustness with Cross Validation

```{r}
f_snaive <- function(y, h) { snaive(y, h = h) }
f_naive  <- function(y, h) { naive(y, h = h) }

# Compute Cross-validation Errors (RMSE)
cv_snaive <- tsCV(ts_load_train, f_snaive, h = 1)
cv_naive  <- tsCV(ts_load_train, f_naive, h = 1)

# Calculate RMSE from CV
rmse_cv_snaive <- sqrt(mean(cv_snaive^2, na.rm = TRUE))
rmse_cv_naive  <- sqrt(mean(cv_naive^2, na.rm = TRUE))

cat("CV RMSE for Seasonal Naive:", rmse_cv_snaive, "\n")
cat("CV RMSE for Naive:", rmse_cv_naive, "\n")


```
```{r}
# 1. Arithmetic Mean Fit
fit_mean <- meanf(ts_load_train, h = n_val)

# 2. Naive Fit
fit_naive <- naive(ts_load_train, h = n_val)

# 3. Seasonal Naive Fit
fit_snaive <- snaive(ts_load_train, h = n_val)

# Compare Accuracy on Validation Set
accuracy_table <- rbind(
  Mean = forecast::accuracy(fit_mean, ts_load_val)[2,],
  Naive = forecast::accuracy(fit_naive, ts_load_val)[2,],
  SNaive = forecast::accuracy(fit_snaive, ts_load_val)[2,]
)
print(accuracy_table)
```


## Generating Final Forecast for Kaggle Submission
```{r}
#Retrain seasonal naive on the full dataset
#ts_daily_load already covers Jan 2005 - Jun 2011
snaive_final <- snaive(ts_daily_load, h = 31)

#Plot the final forecast
autoplot(snaive_final) +
  ylab("Daily Mean Load") +
  ggtitle("Seasonal Naive Forecast: Jul 2011 (Final Submission)")
```
## Creating Submission File

The submission file must match the format of the provided template: a CSV with columns `date` and `load`.
```{r}
#Load submission template
submission_template <- read.csv("/Users/meilishen/Documents/TimeSeries/ShenWu_ENV797_TSA_ForecastCompetition_S26/Ouput/submission_template.csv")

#Create submission with forecasted values
submission <- submission_template

submission$load <- as.numeric(snaive_final$mean)

#Inspect submission
print(submission)

#Save to csv
write.csv(submission, 
          file = "/Users/meilishen/Documents/TimeSeries/ShenWu_ENV797_TSA_ForecastCompetition_S26/Ouput/submission_snaive1.csv", 
          row.names = FALSE)

```


```{r}

# Retrain on full data
final_mean   <- meanf(ts_daily_load, h = 31)
final_naive  <- naive(ts_daily_load, h = 31)
final_snaive <- snaive(ts_daily_load, h = 31)

# Load template
path_template <- "/Users/meilishen/Documents/TimeSeries/ShenWu_ENV797_TSA_ForecastCompetition_S26/Ouput/submission_template.csv"
template <- read.csv(path_template)


save_sub <- function(forecast_obj, filename) {
  sub <- template
  sub$load <- as.numeric(forecast_obj$mean)
  write.csv(sub, file = paste0("/Users/meilishen/Documents/TimeSeries/ShenWu_ENV797_TSA_ForecastCompetition_S26/Ouput/", filename), row.names = FALSE)
}

# Generate the three files
save_sub(final_mean,   "submission_1_mean.csv")
save_sub(final_naive,  "submission_2_naive.csv")
save_sub(final_snaive, "submission_3_snaive.csv")

```

```{r}

df_final_condensed <- daily_data %>%
  left_join(temp_daily_allsites, by = "Date") %>%   
  left_join(rh_daily_allsites,   by = "Date") %>%
  arrange(Date) %>%
  fill(everything(), .direction = "downup")         

train_data <- df_final_condensed %>% filter(Date <= as.Date("2010-06-30"))
val_data   <- df_final_condensed %>% filter(Date >= as.Date("2010-07-01") &
                                             Date <= as.Date("2010-07-31"))
full_data  <- df_final_condensed %>% filter(Date <= as.Date("2011-06-30"))

cat("Split sizes — Train:", nrow(train_data),
    "| Val:", nrow(val_data),
    "| Full:", nrow(full_data), "\n")

historical_julys <- df_final_condensed %>%
  filter(month(Date) == 7, year(Date) < 2011) %>%
  mutate(
    day_of_month = day(Date),
    year_weight  = year(Date) - 2004   # 2005→1, 2006→2, …, 2010→6
  )

july_2011_wtd <- historical_julys %>%
  group_by(day_of_month) %>%
  summarise(across(
    .cols  = where(is.numeric) & !contains("load") & !contains("year_weight"),
    .fns   = ~ weighted.mean(.x, w = year_weight, na.rm = TRUE)
  )) %>%
  ungroup() %>%
  mutate(Date = as.Date(paste0("2011-07-", str_pad(day_of_month, 2, pad = "0")))) %>%
  select(Date, everything(), -day_of_month)

cat("Historic-mean regressor rows:", nrow(july_2011_wtd), "\n")

project_series <- function(col_data, n_ahead = 31) {
  # col_data: named numeric vector with Date names (for ordering)
  ts_obj <- ts(col_data, frequency = 365.25)
  fit    <- auto.arima(ts_obj, seasonal = FALSE, stepwise = TRUE, approximation = TRUE)
  as.numeric(forecast(fit, h = n_ahead)$mean)
}

regressor_cols <- df_final_condensed %>%
  filter(Date <= as.Date("2011-06-30")) %>%
  select(where(is.numeric) & !contains("load")) %>%
  names()

july_dates_2011 <- seq(as.Date("2011-07-01"), as.Date("2011-07-31"), by = "day")

july_2011_model <- map_dfc(regressor_cols, function(col) {
  col_vec <- full_data %>% arrange(Date) %>% pull(!!sym(col))
  preds   <- project_series(col_vec, n_ahead = 31)
  tibble(!!col := preds)
}) %>%
  mutate(Date = july_dates_2011) %>%
  select(Date, everything())

cat("Model-based regressor rows:", nrow(july_2011_model), "\n")


REGRESSOR_APPROACH <- "wtd_mean"   # or "model_based"

july_2011_features <- if (REGRESSOR_APPROACH == "wtd_mean") july_2011_wtd else july_2011_model


resamples_tscv <- time_series_cv(
  data       = train_data,
  date_var   = Date,
  initial    = "3 years",
  assess     = "1 month",
  skip       = "1 month",
  cumulative = TRUE
)

cat("CV slices:", nrow(resamples_tscv), "\n")

rec_prophet <- recipe(daily_mean_load ~ Date + daily_mean_temp + daily_mean_rh,
                      data = train_data)

rec_ml <- recipe(daily_mean_load ~ ., data = train_data) %>%
  step_date(Date, features = c("month", "week", "dow"), label = TRUE) %>%
  step_mutate(Date_num = as.numeric(Date)) %>%
  step_rm(Date) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_nzv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())   # helps GLM & some tree variants

spec_prophet <- prophet_reg(
  seasonality_yearly = TRUE,
  seasonality_weekly = TRUE
) %>% set_engine("prophet")

spec_xgb <- boost_tree(trees = 1000, tree_depth = 3, learn_rate = 0.05) %>%
  set_engine("xgboost", objective = "reg:squarederror") %>%
  set_mode("regression")

spec_arima <- arima_reg() %>% set_engine("auto_arima")

spec_glmnet <- linear_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("regression")

fit_prophet <- workflow() %>%
  add_model(spec_prophet) %>%
  add_recipe(rec_prophet) %>%
  fit(train_data)

fit_xgb <- workflow() %>%
  add_model(spec_xgb) %>%
  add_recipe(rec_ml) %>%
  fit(train_data)

fit_arima <- workflow() %>%
  add_model(spec_arima) %>%
  add_recipe(recipe(daily_mean_load ~ Date + daily_mean_temp + daily_mean_rh,
                    data = train_data)) %>%
  fit(train_data)

cat("Baseline models fitted.\n")

spec_prophet_tune <- prophet_reg(
  seasonality_yearly       = TRUE,
  seasonality_weekly       = TRUE,
  prior_scale_changepoints = tune(),
  prior_scale_seasonality  = tune()
) %>% set_engine("prophet")

prophet_grid <- expand.grid(
  prior_scale_changepoints = c(0.005, 0.01, 0.05, 0.1),
  prior_scale_seasonality  = c(0.1, 0.5, 1,5,10)
)

cat("Tuning Prophet (", nrow(prophet_grid), "x", nrow(resamples_tscv), "folds)...\n")

tune_prophet <- tune_grid(
  workflow() %>% add_model(spec_prophet_tune) %>% add_recipe(rec_prophet),
  resamples = resamples_tscv,
  grid      = prophet_grid,
  control   = control_grid(verbose = FALSE, save_pred = TRUE)
)

best_prophet <- select_best(tune_prophet, metric = "rmse")
cat("Best Prophet params:\n"); print(best_prophet)

fit_prophet_opt <- finalize_workflow(
  workflow() %>% add_model(spec_prophet_tune) %>% add_recipe(rec_prophet),
  best_prophet
) %>% fit(train_data)


spec_xgb_tune <- boost_tree(
  trees      = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  min_n      = tune()
) %>%
  set_engine("xgboost", objective = "reg:squarederror") %>%
  set_mode("regression")

xgb_grid <- expand.grid(
  trees      = c(500),
  tree_depth = c(3),
  learn_rate = c(0.01, 0.05),
  min_n      = c(5, 20)
)

cat("Tuning XGBoost (", nrow(xgb_grid), "combinations)...\n")

tune_xgb <- tune_grid(
  workflow() %>% add_model(spec_xgb_tune) %>% add_recipe(rec_ml),
  resamples = resamples_tscv,
  grid      = xgb_grid,
  control   = control_grid(verbose = FALSE)
)

best_xgb <- select_best(tune_xgb, metric = "rmse")
cat("Best XGBoost params:\n"); print(best_xgb)

fit_xgb_opt <- finalize_workflow(
  workflow() %>% add_model(spec_xgb_tune) %>% add_recipe(rec_ml),
  best_xgb
) %>% fit(train_data)


glmnet_grid <- expand.grid(
  penalty  = 10^seq(-4, 0, length.out = 6),
  mixture  = c(0, 0.25, 0.5, 0.75, 1)    # 0=Ridge, 1=Lasso
)

cat("Tuning GLMNet (", nrow(glmnet_grid), "combinations)...\n")

tune_glmnet <- tune_grid(
  workflow() %>% add_model(spec_glmnet) %>% add_recipe(rec_ml),
  resamples = resamples_tscv,
  grid      = glmnet_grid,
  control   = control_grid(verbose = FALSE)
)

best_glmnet <- select_best(tune_glmnet, metric = "rmse")
cat("Best GLMNet params:\n"); print(best_glmnet)
fit_glmnet_opt <- finalize_workflow(
  workflow() %>% add_model(spec_glmnet) %>% add_recipe(rec_ml),
  best_glmnet
) %>% fit(train_data)


model_table <- modeltime_table(
  fit_prophet,      # M1: Prophet baseline
  fit_prophet_opt,  # M2: Prophet tuned
  fit_xgb,          # M3: XGBoost baseline
  fit_xgb_opt,      # M4: XGBoost tuned
  fit_arima,        # M5: Auto-ARIMA
  fit_glmnet_opt    # M6: GLMNet tuned
)

calibration_table <- model_table %>%
  modeltime_calibrate(new_data = val_data)

cat("\n── Validation Accuracy (July 2010 holdout) ──\n")
print(modeltime_accuracy(calibration_table))


cat("\nRefitting all models on full dataset (Jan 2005 – Jun 2011)...\n")
refit_table <- calibration_table %>%
  modeltime_refit(data = full_data)


final_fcst <- refit_table %>%
  modeltime_forecast(
    new_data    = july_2011_features,
    actual_data = full_data
  )

preds_list <- list(
  prophet_base  = final_fcst %>% filter(.model_id == 1, .key == "prediction") %>% pull(.value),
  prophet_tuned = final_fcst %>% filter(.model_id == 2, .key == "prediction") %>% pull(.value),
  xgb_base      = final_fcst %>% filter(.model_id == 3, .key == "prediction") %>% pull(.value),
  xgb_tuned     = final_fcst %>% filter(.model_id == 4, .key == "prediction") %>% pull(.value),
  arima         = final_fcst %>% filter(.model_id == 5, .key == "prediction") %>% pull(.value),
  glmnet        = final_fcst %>% filter(.model_id == 6, .key == "prediction") %>% pull(.value)
)


# ── CHUNK 14: ENSEMBLES ───────────────────────────────────────────────────────
# Three ensemble strategies. Accuracy-weighted uses val RMSE as inverse weights.

ensemble_simple <- rowMeans(bind_cols(preds_list))

val_rmse <- modeltime_accuracy(calibration_table) %>%
  arrange(.model_id) %>%
  pull(rmse)

inv_rmse_weights <- (1 / val_rmse) / sum(1 / val_rmse)
cat("Ensemble weights (inv-RMSE):", round(inv_rmse_weights, 3), "\n")

ensemble_wtd <- as.matrix(bind_cols(preds_list)) %*% inv_rmse_weights %>% as.numeric()

top2_ids <- order(val_rmse)[1:2]
ensemble_top2 <- rowMeans(bind_cols(preds_list)[, top2_ids])

cat("\nEnsemble preview (first 5 days):\n")
print(tibble(
  Date            = july_dates_2011[1:5],
  simple_avg      = round(ensemble_simple[1:5]),
  weighted_avg    = round(ensemble_wtd[1:5]),
  top2_avg        = round(ensemble_top2[1:5])
))


save_sub(list(mean = preds_list$prophet_base),  "Sub_Prophet_Base.csv")
save_sub(list(mean = preds_list$prophet_tuned), "Sub_Prophet_Tuned.csv")
save_sub(list(mean = preds_list$xgb_base),      "Sub_XGBoost_Base.csv")
save_sub(list(mean = preds_list$xgb_tuned),     "Sub_XGBoost_Tuned.csv")
save_sub(list(mean = preds_list$arima),          "Sub_ARIMA.csv")
save_sub(list(mean = preds_list$glmnet),         "Sub_GLMNet.csv")
save_sub(list(mean = ensemble_simple),           "Sub_Ensemble_Simple.csv")
save_sub(list(mean = ensemble_wtd),              "Sub_Ensemble_Weighted.csv")
save_sub(list(mean = ensemble_top2),             "Sub_Ensemble_Top2.csv")

cat("\nAll submissions saved.\n")
```
